Characteristic polynomials of p-adic matrices

### 2017-04-15 charpoly/charpoly.tex
Notation for precision uniformized

\begin{abstract}
We analyze the precision of the characteristic polynomial $\chi(A)$ of
an $n \times n$ $p$-adic matrix $A$ using differential precision methods
developed previously.  When $A$ is integral with precision $O(p^k)$,
we give a criterion (checkable in time $\softO(n^\omega)$) for
$\chi(A)$ to have precision exactly $O(p^k)$.  We also give a $\softO(n^3)$
algorithm for determining the optimal precision when the criterion is not
satisfied, and give examples when the precision is larger than $O(p^k)$.
\end{abstract}

@@

\vspace{-1.5mm}
\keywords{Algorithms, $p$-adic precision, Newton polygon, factorization}

@@

When computing with $p$-adic matrices, the lack of infinite memory implies
that the entries may only be approximated at some finite precision $O(p^m)$.  As a
consequence, in designing algorithms for such matrices one must analyze
not only the running time of the algorithm but also the accuracy of the result.

Let $M \in M_n(\Qp)$ be known at precision $O(p^m)$.
The simplest approach for computing the characteristic polynomial of $M$
is to compute $\det(X-M)$ either using recursive row expansion or various division free
algorithms \citelist{\cite{seifullin:02a} \cite{kaltofen:92a}}.  There are two issues
with these methods.  First, they are slower than alternatives that allow division,
requiring $O(n!)$, $O(n^4)$ and $\softO(n^{2 + \omega/2})$ operations.  Second,
while the lack of division implies that the result is accurate modulo $p^m$ as
long as $M \in M_n(\Zp)$, they still do not yield the optimal precision.

@@

One benefit of this approach is that we may account for diffuse precision:
precision that is not localized on any single coefficient of $\chi_M$.  For example,
let $0 \le \alpha_1 \le \alpha_2 \le \dots \le \alpha_n$, consider a
diagonal matrix $D$ with diagonal entries $(p^{\alpha_1}, \dots, p^{\alpha_n})$,
let $P, Q \in \GL_n(\Zp)$ and set $M = PDQ$.  The valuation of the coefficient
of $X^{n-k}$ in $\chi_M$ will be $\sum_{i=1}^k \alpha_i$, and if $\alpha_{n-1} > 0$
and $M$ is known with precision $O(p^m)$ then the constant term of $\chi_M$
will be known with precision larger than $O(p^m)$ (see
\cite{caruso-roe-vaccon:15a}*{Prop. 3.2}).

As long as none of the eigenvalues of $M$ are congruent to $-1$ modulo $p$,
then none of the coefficients of the characteristic polynomial of $1+M$ will have
precision larger than $O(p^m)$.  But $\chi_{1+M}(X) = \chi_M(X-1)$, so the precision
content of these two polynomials should be equivalent. The solution is that the
extra precision in $\chi_{1+M}$ is diffuse and not visible on any individual coefficient.
We formalize this phenomenon using lattices; see Section \ref{sec:padicprec} for
further explanation, and \cite{caruso:17a}*{\S 3.2.2} for a specific example of the
relationship between $\chi_M$ and $\chi_{1+M}$.

@@

We recall some of the definitions and results of \cite{caruso-roe-vaccon:14a}
as a foundation for our discussion of the precision for the characteristic polynomial
of a matrix.  We will be concerned with two $K$-manifolds in what follows:
the space $M_n(K)$ of $n \times n$ matrices with entries in $K$ and the space
$K_n[X]$ of monic degree $n$ polynomials over $K$.  Given a matrix $M \in M_n(K)$,
the most general kind of precision structure we may attach to $M$ is a
\emph{lattice} $H$ in the tangent space at $M$.  However, representing an
arbitrary lattice requires $n^2$ basis vectors, each with $n^2$ entries.  We therefore
frequently work with certain classes of lattices, either \emph{jagged} lattices
where we specify a precision for each matrix entry or \emph{flat} lattices where
every entry is known to a fixed precision $O(p^m)$.  Similarly, precision for
monic polynomials can be specified by giving a lattice in the tangent space
at $f(X) \in K_n[X]$, or restricted to jagged or flat precision in the interest
of simplicity.

>>>>

\begin{abstract}
We analyze the precision of the characteristic polynomial $\chi(A)$ of
an $n \times n$ $p$-adic matrix $A$ using differential precision methods
developed previously.  When $A$ is integral with precision $O(p^N)$,
we give a criterion (checkable in time $\softO(n^\omega)$) for
$\chi(A)$ to have precision exactly $O(p^N)$.  We also give a $\softO(n^3)$
algorithm for determining the optimal precision when the criterion is not
satisfied, and give examples when the precision is larger than $O(p^N)$.
\end{abstract}

@@

\vspace{-1.5mm}
\keywords{Algorithms, $p$-adic precision, characteristic polynomial,
eigenvalue}

@@

When computing with $p$-adic matrices, the lack of infinite memory implies
that the entries may only be approximated at some finite precision $O(p^N)$.  As a
consequence, in designing algorithms for such matrices one must analyze
not only the running time of the algorithm but also the accuracy of the result.

Let $M \in M_n(\Qp)$ be known at precision $O(p^N)$.
The simplest approach for computing the characteristic polynomial of $M$
is to compute $\det(X-M)$ either using recursive row expansion or various division free
algorithms \citelist{\cite{seifullin:02a} \cite{kaltofen:92a}}.  There are two issues
with these methods.  First, they are slower than alternatives that allow division,
requiring $O(n!)$, $O(n^4)$ and $\softO(n^{2 + \omega/2})$ operations.  Second,
while the lack of division implies that the result is accurate modulo $p^N$ as
long as $M \in M_n(\Zp)$, they still do not yield the optimal precision.

@@

One benefit of this approach is that we may account for diffuse precision:
precision that is not localized on any single coefficient of $\chi_M$.  For example,
let $0 \le \alpha_1 \le \alpha_2 \le \dots \le \alpha_n$, consider a
diagonal matrix $D$ with diagonal entries $(p^{\alpha_1}, \dots, p^{\alpha_n})$,
let $P, Q \in \GL_n(\Zp)$ and set $M = PDQ$.  The valuation of the coefficient
of $X^{n-k}$ in $\chi_M$ will be $\sum_{i=1}^k \alpha_i$, and if $\alpha_{n-1} > 0$
and $M$ is known with precision $O(p^N)$ then the constant term of $\chi_M$
will be known with precision larger than $O(p^N)$ (see
\cite{caruso-roe-vaccon:15a}*{Prop. 3.2}).

As long as none of the eigenvalues of $M$ are congruent to $-1$ modulo $p$,
then none of the coefficients of the characteristic polynomial of $1+M$ will have
precision larger than $O(p^N)$.  But $\chi_{1+M}(X) = \chi_M(X-1)$, so the precision
content of these two polynomials should be equivalent. The solution is that the
extra precision in $\chi_{1+M}$ is diffuse and not visible on any individual coefficient.
We formalize this phenomenon using lattices; see Section \ref{sec:padicprec} for
further explanation, and \cite{caruso:17a}*{\S 3.2.2} for a specific example of the
relationship between $\chi_M$ and $\chi_{1+M}$.

@@

We recall some of the definitions and results of \cite{caruso-roe-vaccon:14a}
as a foundation for our discussion of the precision for the characteristic polynomial
of a matrix.  We will be concerned with two $K$-manifolds in what follows:
the space $M_n(K)$ of $n \times n$ matrices with entries in $K$ and the space
$K_n[X]$ of monic degree $n$ polynomials over $K$.  Given a matrix $M \in M_n(K)$,
the most general kind of precision structure we may attach to $M$ is a
\emph{lattice} $H$ in the tangent space at $M$.  However, representing an
arbitrary lattice requires $n^2$ basis vectors, each with $n^2$ entries.  We therefore
frequently work with certain classes of lattices, either \emph{jagged} lattices
where we specify a precision for each matrix entry or \emph{flat} lattices where
every entry is known to a fixed precision $O(p^N)$.  Similarly, precision for
monic polynomials can be specified by giving a lattice in the tangent space
at $f(X) \in K_n[X]$, or restricted to jagged or flat precision in the interest
of simplicity.
